---
title: "HW 01"
author: "Nilotpal Rajbongshi"
format: 
  html:
    embed-resources: true
toc: true
---

## 0 - Setup

```{r setup}
if (!require("pacman")) 
  install.packages("pacman")

# use this line for installing/loading
pacman::p_load(tidyverse, 
               lubridate, 
               devtools, 
               openintro,
               ggmap,
               pander) 

devtools::install_github("tidyverse/dsbox")
```

## 1 - Road traffic accidents in Edinburgh

```{r}
library(tidyverse)
library(lubridate)
accidents <- read_csv("data/accidents.csv")

accidents <- accidents %>%
  mutate(
    date_time = dmy_hms(paste(date, as.character(time))),
    time_of_day = hour(date_time) + minute(date_time) / 60,
    day_type = if_else(wday(date_time) %in% c(1, 7), "Weekend", "Weekday"),
    severity = factor(severity, levels = c("Fatal", "Serious", "Slight"))
  )
ggplot(accidents, aes(x = time_of_day, fill = severity)) +
  geom_density(alpha = 0.6) +
  facet_wrap(~ day_type, ncol = 1) +
  scale_fill_manual(values = c("Fatal" = "#9970AB", "Serious" = "#5AAE61", "Slight" = "#FFFC5F")) +
  labs(
    title = "Number of accidents throughout the day",
    subtitle = "By day of week and severity",
    x = "Time of day",
    y = "Density",
    fill = "Severity"
  ) +
  theme_minimal()
```

## 2 - NYC marathon winners

```{r}
library(tidyverse)

# Load the CSV file
nyc_marathon <- read_csv("data/nyc_marathon.csv")

# Clean the data
nyc_marathon <- nyc_marathon %>%
  rename(gender = division)

# Filter out rows with missing time_hrs values
nyc_marathon_clean <- nyc_marathon %>%
  filter(!is.na(time_hrs))

# a. Histogram and boxplot of all marathon times
ggplot(nyc_marathon_clean, aes(x = time_hrs)) +
  geom_histogram(binwidth = 0.1, fill = "steelblue", color = "white") +
  labs(title = "Histogram of NYC Marathon Winning Times", x = "Time (hours)", y = "Count")

ggplot(nyc_marathon_clean, aes(y = time_hrs)) +
  geom_boxplot(fill = "tomato") +
  labs(title = "Boxplot of NYC Marathon Winning Times", y = "Time (hours)")

# b. Side-by-side boxplots by gender
ggplot(nyc_marathon_clean, aes(x = gender, y = time_hrs, fill = gender)) +
  geom_boxplot() +
  scale_fill_manual(values = c("Men" = "lightblue", "Women" = "pink")) +
  labs(title = "Marathon Times by Gender", x = "Gender", y = "Time (hours)")

# c. Remove redundant color legend
ggplot(nyc_marathon_clean, aes(x = gender, y = time_hrs, fill = gender)) +
  geom_boxplot(show.legend = FALSE) +
  scale_fill_manual(values = c("Men" = "lightblue", "Women" = "pink")) +
  labs(title = "Marathon Times by Gender (Simplified)", x = "Gender", y = "Time (hours)")

# d. Time series of marathon times by year and gender
ggplot(nyc_marathon_clean, aes(x = year, y = time_hrs, color = gender, shape = gender)) +
  geom_point(size = 3) +
  geom_line(aes(group = gender), linetype = "dashed") +
  scale_color_manual(values = c("Men" = "lightblue", "Women" = "pink")) +
  labs(title = "NYC Marathon Winning Times Over Years",
       x = "Year", y = "Time (hours)", color = "Gender", shape = "Gender")
```

## 3 - US counties
a. What does the following code do? Does it work? Does it make sense? Why/why not?
```{r label-me-3}
ggplot(county) +
  geom_point(aes(x = median_edu, y = median_hh_income)) +
  geom_boxplot(aes(x = smoking_ban, y = pop2017))
```
Here the code is trying to plot a scatter plot of median_edu vs median_hh_income and add a box plot showing distribution of pop2017 across smoking ban. The does run without error, but does not provide the desired plotting result. It does not make sense since we are trying to mix two different x axis variables one is categorical and one is numerical.It cannot mix two different variables across variables.

b. Which of the following two plots makes it easier to compare poverty levels (poverty) across people from different median education levels (median_edu)? What does this say about when to place a faceting variable across rows or columns?
```{r }
ggplot(county %>% filter(!is.na(median_edu))) + 
  geom_point(aes(x = homeownership, y = poverty)) + 
  facet_grid(median_edu ~ .)
ggplot(county %>% filter(!is.na(median_edu))) + 
  geom_point(aes(x = homeownership, y = poverty)) + 
  facet_grid(. ~ median_edu)
```
Here the second plot makes it lot more easier to compare poverty levels across people from different median education levels as poverty on y axis aligns across columns which allows direct visual comparison, whereas poverty y axis in the first plot is not aligned side by side. Also its easier for the eye to scan the data point top to bottom. If the values are to be compared across y groups use horizontal facets. If across x axis use vertical facets.

c. Recreate the R code necessary to generate the following graphs. Note that wherever a categorical variable is used in the plot, it’s metro.
```{R}
library(openintro)
data(county)
#Plot A
ggplot(county, aes(x = homeownership, y = poverty)) + 
  geom_point() + labs(title = "Plot A")
```
```{R}
#Plot B
ggplot(county, aes(x = homeownership, y = poverty)) +
  geom_point() +
  geom_smooth(color = "blue", se = FALSE) +labs(title = "Plot B")

```
```{R}
ggplot(county, aes(x = homeownership, y = poverty)) +
  geom_point() +
  stat_smooth(method = "", formula = y ~ s(x), se = FALSE, aes(group = metro), color = "green") +
  ggtitle("Plot C")
```
```{R}
ggplot(county, aes(x = homeownership, y = poverty)) +
  stat_smooth(method = "gam", formula = y ~ s(x), se = FALSE, aes(group = metro), color = "BLUE")+
  geom_point() + 
  labs(title = "Plot D")
```
```{R}
ggplot(county, aes(x = homeownership, y = poverty, color = metro, linetype = metro)) +
  geom_point() +
  geom_smooth(method = "gam", formula = y ~ s(x), se = FALSE, aes(group = metro), color = "BLUE") +
  labs(title = "Plot E")
```
```{R}
ggplot(county, aes(x = homeownership, y = poverty, color = metro, linetype = metro)) +
  geom_point() +
  geom_smooth(se = FALSE) +
  scale_linetype_manual(values = c("no" = "solid", "yes" = "solid")) +
  labs(title = "Plot F")
```
```{R}
ggplot(county, aes(x = homeownership, y = poverty)) +
  geom_point(aes(color = metro)) +
  geom_smooth(color = "blue", se = FALSE) +labs(title = "Plot G")
```
```{R}
ggplot(county, aes(x = homeownership, y = poverty, color = metro)) +
  geom_point() +labs(title = "Plot H")

```

## 4 - Rental apartments in SF
```{R ques a}
library(tidyverse)

# Load and clean the data
credit <- read_csv("data/credit.csv") %>%
  rename_with(tolower) %>%
  mutate(
    student = str_trim(str_to_title(student)),
    married = str_trim(str_to_title(married)),
    student = factor(student, levels = c("No", "Yes")),
    married = factor(married, levels = c("No", "Yes"))
  )

# Create the exact plot
ggplot(credit, aes(x = income, y = balance, color = student, shape = student)) +
  geom_point(alpha = 0.6, size = 2) +
  geom_smooth(method = "lm", se = FALSE, size = 1) +
  facet_grid(student ~ married, labeller = label_both) +
  scale_color_manual(values = c("No" = "#D6B85A", "Yes" = "#2E6F40")) + 
  scale_shape_manual(values = c("No" = 16, "Yes" = 17)) +
  scale_x_continuous(labels = scales::dollar_format(prefix = "$", suffix = "K")) +
  scale_y_continuous(labels = scales::dollar_format(prefix = "$")) +
  labs(
    x = "Income",
    y = "Credit card balance"
  ) +
  theme_minimal(base_size = 13) +
  theme(
    legend.position = "none",
    strip.background = element_rect(fill = "grey90", color = NA),
    strip.text = element_text(face = "bold"),
    panel.border = element_rect(color = "grey30", fill = NA),
    panel.spacing = unit(1, "lines")
  )

```
b. Yes, married and student status could be useful predictors for credit card balance alongside income. The visualization suggests students tend to have higher balances, and marital status influences spending patterns. Including these variables in a model could improve balance predictions by accounting for these behavioral differences.
```{R ques c}
credit <- credit %>%
  mutate(utilization = balance / limit)
ggplot(credit, aes(x = income, y = utilization, color = student, shape = student)) +
  geom_point(alpha = 0.6, size = 2) +
  geom_smooth(method = "lm", se = FALSE, size = 1) +
  facet_grid(student ~ married, labeller = label_both) +
  scale_color_manual(values = c("No" = "#D6B85A", "Yes" = "#2E6F40")) + 
  scale_shape_manual(values = c("No" = 16, "Yes" = 17)) +
  labs(
    x = "Income",
    y = "Credit Utilization"
  ) +
  theme_minimal(base_size = 13) +
  theme(
    legend.position = "none",
    strip.background = element_rect(fill = "grey90", color = NA),
    strip.text = element_text(face = "bold"),
    panel.border = element_rect(color = "grey30", fill = NA),
    panel.spacing = unit(1, "lines")
  )
```
d.Income and credit utilization show a different pattern than income and credit balance. Higher-income individuals tend to have lower credit utilization, while lower-income individuals rely more on their available credit. Student and married status also influence utilization rates, reflecting borrowing habits beyond just spending levels.

## 5 - Napoleon’s march.
```{R}
# ------------------------------------------------------------------
# Resources Used
# ------------------------------------------------------------------
# 1. Exploring Minard’s 1812 plot with #ggplot2:https://www.andrewheiss.com/blog/2017/08/10/exploring-minards-1812-plot-with-ggplot2/
#
# 2. Minard's Map context – https://www.edwardtufte.com/tufte/minard
#    Referred to the original visualization to guide plot design choices.
# ------------------------------------------------------------------

#Loads necessary R packages for data manipulation, visualization, and formatting.
library(tidyverse)
library(lubridate)
library(ggmap)
library(ggrepel)
library(gridExtra)
library(pander)

#read_rds reads R data file containing Minard's Napolean dataset
napoleon <- read_rds("data/napoleon.rds")

#the data includes cities, troops, temps
cities <- napoleon$cities
troops <- napoleon$troops
temps <- napoleon$temperatures

#Below code draws the basic movement paths of the troops from west to east and back.
#group=group ensures correct path segments are drawn
ggplot(troops, aes(x = long, y = lat, group = group)) +
  geom_path()

#Enhances the plot by adding aesthetics. Adding color and size.
ggplot(troops, aes(x = long, y = lat, group = group,
                   color = direction, size = survivors)) +
  geom_path()

#Makes line ends rounded instead of square, improving visual smoothness. Enhances the aesthetic quality of the paths.
ggplot(troops, aes(x = long, y = lat, group = group,
                   color = direction, size = survivors)) +
  geom_path(lineend = "round")

 #Sets the minimum and maximum line widths for survivors, making the visualization more dramatic.Ensures line thickness clearly reflects the number of survivors.
ggplot(troops, aes(x = long, y = lat, group = group,
                   color = direction, size = survivors)) +
  geom_path(lineend = "round") +
  scale_size(range = c(0.5, 15))

#scale_colour_manual sets custom colors for direction ,labs() Removes axis labels for a minimalist look. guides(), hides legends for color and size, mimicking Minard’s clean style.

ggplot(troops, aes(x = long, y = lat, group = group,
                   color = direction, size = survivors)) +
  geom_path(lineend = "round") +
  scale_size(range = c(0.5, 15)) +
  scale_colour_manual(values = c("#808080", "#252523")) +
  labs(x = NULL, y = NULL) +
  guides(color = FALSE, size = FALSE)

#Adds city locations and labels to the troop plot.geom_point(), plots cities as points at their coordinates.
ggplot() +
  geom_path(data = troops, aes(x = long, y = lat, group = group,
                               color = direction, size = survivors),
            lineend = "round") +
  geom_point(data = cities, aes(x = long, y = lat)) +
  geom_text(data = cities, aes(x = long, y = lat, label = city), vjust = 1.5) +
  scale_size(range = c(0.5, 15)) +
  scale_colour_manual(values = c("#808080", "#252523")) +
  labs(x = NULL, y = NULL) +
  guides(color = FALSE, size = FALSE)

#geom_point(), Colors city points red for visibility.geom_text_repel(), uses ggrepel to place non-overlapping city labels, with red text and a specific font.Creates a clearer, more professional map with readable city labels.
ggplot() +
  geom_path(data = troops, aes(x = long, y = lat, group = group,
                               color = direction, size = survivors),
            lineend = "round") +
  geom_point(data = cities, aes(x = long, y = lat),
             color = "#0000FF") +
  geom_text_repel(data = cities, aes(x = long, y = lat, label = city),
                  color = "#0000FF", family = "Open Sans Condensed Bold") +
  scale_size(range = c(0.5, 15)) +
  scale_colour_manual(values = c("#808080", "#252523")) +
  labs(x = NULL, y = NULL) +
  guides(color = FALSE, size = FALSE)

```
